Estimating Real-Time Highstreet Footfall from Wi-Fi Probe Requests

### 2018-12-07 paper/sections/methodology.tex
methodology corrections

The primary aim of this research is to enable us to collect a series of probe
requests and process them into a usable pedestrian footfall count. We do this by
using a Wi-Fi receiver to collect probe requests broadcast by mobile devices,
filtering out the background noise and aggregating them based on the device that
generated them. In this section, we begin by looking at the characteristics of
probe requests in detail, device a methodology to collect these probe requests
in public areas, examine the systemic biases and uncertainties in the data
collection method and device data processing methods to overcome these
challenges. Finally we compare the processed footfall counts to the ground truth
recorded by primary surveys.

Probe requests are a special type of management packets broadcast by Wi-Fi
enabled devices as part of the various functions such as scanning for available
APs, quick geolocation by triangulation based known APs, etc. These are
broadcast by all Wi-Fi enabled devices regardless of the manufacturer, type or
model of the devices though there is some variation on the frequency and the
information transmitted through them. In some cases, such as Android devices,
these are broadcast even when the Wi-Fi functionality has been turned off by
the user so that device can immediately connet to networks when the
functionality is switched back on. Since some devices even use the probe
requests as a less accurate form of localisation they keep sending probe
requests as well. Thus these signals can be used to reliably identify the
presence of Wi-Fi enabled mobile devices. Being a first step of connection
initiated by the mobile device, these packets have information regarding the
characteristics of the mobile device itself. Some of the key information we can
infer from these requests are,

\begin{enumerate} 
\item \textbf{Media Access Control (MAC) address} which is an name identifying
	for the wireless hardware of the mobile device, 
\item \textbf{Sequence number} of the request for the mobile device to keep
	track of the responses, 
\item \textbf{Time stamp} at which the request was received by the AP, 
\item Total \textbf{length} of the request in number of bits, and 
\item The \textbf{strength of the signal} received by the mobile device.
\end{enumerate}

The MAC address is the primary unique identifier for the mobile device. It has
two parts, first part is the Organisationally Unique Identifier (OUI) which
gives information about the manufacturer of the device and the second is unique
to the device. In modern devices, to protect the users privacy, the MAC address
can also be randomised (hence non unique) and is marked as such. Though
sequence number of the packet is strictly unique to a mobile device, we
hypothesize that we can use them to estimate the number unique devices as
demonstrated by \citep{vanhoef2016}, where optional information present in the
probe requests - Information Elements (IE) along with the probe requests have
been used to fingerprint the devices. This approach has become increasinly
difficult as mobile phone manufacturers have severely limited the IEs present
in the probe requests thus leading us to explore methods which uses only the
sequence numbers. This also affects the established commercial solutions using
Wi-Fi probe requests such as Blix, Walkbase, Euclid Analytics, RetailNext etc.
There have been another solution proposed by \citep{hong2018crowdprobe} where
the authors tried to solve the similar problem using a hidden markov models
based trajectory inference alogorithm but the scope is limited to enclosed,
exit controlled public spaces such as shopping malls, railway stations, etc.

Data collection was done with the help of custom sensors built from modifying
the hardware used in Smart street sensors \citep{sss2016} and updating them with
custom software. The sensor is essentially a Raspberry-Pi device with Wi-Fi and
3G modules. It keeps the Wi-Fi module in `Monitor' mode and uses the open source
software - Wireshark \citep{wireshark2} to passively collect all packets sent to
`broadcast', marked with type as management' and subtype as `probe requests'.
The MAC address in these probe requests is obfuscated at the device level using
a cryptographic hashing algorithm and transmitted through 3G connection to a
central database via web-sockets protocol, where it is stored in a PostgreSQL
database for further analysis. The random salt used in the hashing algorithm was
rotated regularly to further mitigate the risk of de-anonymisation of the hash.
Though hashing cannot completely ensure anonymisation as discussed in
\citep{demir2014analysing} it can sufficiently obfuscate the data which along
with a secure process of data handling gives us reasonable security. A overall
schematic of the data collection and storage process is shown in Figure
\ref{datacollection_schematic}. The ground truth on number of pedestrian
footfall was recorded using a custom Android application - Clicker
\citep{bala2018clicker}. This app logs accurate timestamps for each time the
surveryor records a pedestrian crossing the designated cordon line at the
location. In addition to counting the pedestrians manually this also makes the
mobile device used constantly probe for networks which also gives us a know
device to calibrate our methodology against.

@@

The next step after collecting data was to estimate the footfall or pedestrian
activity from them. We identified the following potential uncertainties which
arise from our collection methodology.

>>>>

The primary aim of this research was to enable us to collect a series of probe
requests and process them into a usable pedestrian footfall count. We did this
by using a Wi-Fi receiver to collect probe requests broadcast by mobile
devices, filtering out the background noise, and aggregating them based on the
device that generated them. In this section, we examine the
characteristics of probe requests in detail, devise a methodology to collect
these probe requests in public areas, examine the systemic biases and
uncertainties in the data collection method, and devise data processing methods
to overcome these challenges. Finally, we compare the processed footfall counts
to the ground truth recorded by primary surveys.

Probe requests are a special type of management packet broadcast by Wi-Fi
enabled devices as part of their various functions such as scanning for
available APs and quick geolocation by triangulation based known APs, etc,.
These are broadcast by all Wi-Fi enabled devices regardless of the manufacturer,
type or model of the devices, although there is some variation in the frequency
and the content of the information transmitted through them. In some cases, such
as Android devices, these are broadcast even when the Wi-Fi functionality has
been turned off by the user so that the device can immediately connect to
networks when the functionality is switched back on. Since some devices even use
the probe requests as a less accurate form of localisation, they continuously
send probe requests when Wi-Fi has been switched off. Thus, these signals can be
used to reliably identify the presence of Wi-Fi enabled mobile devices. Being a
first step of connection initiated by the mobile device, these packets have
information regarding the characteristics of the mobile device itself. Some of
the key information we can infer from these requests are,

\begin{enumerate} 
\item \textbf{Media Access Control (MAC) address} which is a name identifying
	the wireless hardware of the mobile device, 
\item \textbf{Sequence number} of the request for the mobile device to keep
	track of the responses, 
\item \textbf{Time stamp} at which the request was received by the AP, 
\item Total \textbf{length} of the request in number of bits, and 
\item The \textbf{strength of the signal} received by the mobile device.
\end{enumerate}

The MAC address is the primary identifier for the mobile device and has two
parts. The first part is the Organisationally Unique Identifier (OUI) which
provides information about the manufacturer of the device and the second part is
the identifier for the device. In modern devices, to protect users' privacy, the
second part of the MAC address can also be randomised and hence may not be
unique to devices. When the MAC address is randomised, it is marked as such by
setting a specific bit in the probe request packet as 1. Although sequence
number of the packet is strictly unique to a mobile device, we hypothesize that
we can use them to estimate the number of unique devices as demonstrated by
\citep{vanhoef2016}; where optional information present in the probe requests -
Information Elements (IE) along with the sequence numbers, have been used to
fingerprint the devices. This approach has become increasingly difficult as
mobile phone manufacturers have severely limited the IEs present in the probe
requests thus leading us to explore methods which use only the sequence numbers.
This also affects the established commercial solutions using Wi-Fi probe
requests such as Blix, Walkbase, Euclid Analytics, RetailNext etc.  There has
been another solution proposed by \citep{hong2018crowdprobe} where the authors
tried to solve the similar problem using a hidden markov models based trajectory
inference algorithm but the scope of this research was limited to enclosed,
exit controlled public spaces such as shopping malls, railway stations, etc.

Data collection was done with the help of custom sensors built from modifying
the hardware used in Smart Street Sensors \citep{sss2016} and updating them with
custom software. The sensor is essentially a Raspberry-Pi device with Wi-Fi and
3G modules. It keeps the Wi-Fi module in `Monitor mode' and uses the open source
software - Wireshark \citep{wireshark2} to passively collect all packets sent to
`broadcast', marked with type as `management', and subtype `probe requests'.
The MAC address in these probe requests is obfuscated at the device level using
a cryptographic hashing algorithm and transmitted through 3G connection to a
central database via web-sockets protocol, where it is stored in a PostgreSQL
database for further analysis. The random salt used in the hashing algorithm was
rotated regularly to further mitigate the risk of de-anonymisation of the hash.
Though hashing cannot completely ensure anonymisation as discussed in
\citep{demir2014analysing}, it can sufficiently obfuscate the data; which along
with a secure process of data handling, gives us reasonable security. An overall
schematic of the data collection and storage process is shown in Figure
\ref{datacollection_schematic}. The ground truth on the number of pedestrian
footfall was recorded using a custom Android application - Clicker
\citep{bala2018clicker}. This app logs accurate timestamps each time the
surveryor records a pedestrian crossing the designated cordon line at the
location. In addition to counting the pedestrians manually, this procedure
results in the device broadcasting probe requests regularly, which in turn,
gives us a `known device' to calibrate our methodology against.

@@

After collecting data, we began estimating the footfall or pedestrian
activity from them by identifying the following potential uncertainties arising
from our data collection method.
